Linting Linearity in Core/System FC

### 2023-01-23 proposal.tex
More on linear haskell

In this sense, Core is already annotated with linearity, but the \textbf{linter currently
  ignores linearity annotations}.
%
In spite of the strong formal foundations of linear types driving the
implementation, their interaction with the whole of GHC is still far
from trivial. The implemented type system cannot accomodate
several optimising transformations that produce programs that violate
linearity syntactically (i.e.,~multiple occurrences of linear
variables in the program text), but ultimately preserve it in a
semantic sense, where a linear term is still consumed exactly once --
this is compounded by lazy evaluation driving a non-trivial mismatch
between syntactic and semantic linearity. Therefore, linear linting
rejects various valid programs (with regard to linearity) after
desugaring. The current solution to this is to effectively disable the
linear linter, since otherwise disabling optimisations can incur significant
performance costs. However, we believe that GHC's transformations are
correct, and it is the linear linter and its underlying type system
that cannot sufficiently accommodate the
resulting programs.
% diverse problems with linearity spring up when we get past the desugarer and
% the
% linter rejects valid programs which is quite undesirable.

@@

Despite not being accepted by the surface-level language, linear programs using
lets occur naturally in Core due to optimising transformations that create let
bindings. In a similar fashion, programs which violate syntatic linearity
for other reasons other than let bindings are produced by Core transformations.

@@

The key idea to computing the usage environment of multiple variables that use
themselves in their definition body is to perform the computation in two separate passes. First,
calculate a naive environment by emiting free variables multiplicities as usual
while treating the recursive-let bound variables as free ones. Second, pass the
variables names and their corresponding naive environment as input to
algorithm~\ref{computeRecUsages} to get a final usage environment. Intuitively,
the algorithm, for each recursive binder, iterates over all (initially naive) usage
environments and substitutes the recursive binder by the corresponding usage
environment, scaled up by the amount of times that recursive binder is used in
the environment being updated.

>>>>

In this sense, Core is already annotated with linearity, but the \textbf{linter currently
  ignores linearity annotations}.
%
In spite of the strong formal foundations of linear types driving the
implementation, their interaction with the whole of GHC is still far
from trivial. The implemented type system cannot accommodate 
several optimising transformations that produce programs that violate
linearity syntactically (i.e.,~multiple occurrences of linear
variables in the program text), but ultimately preserve it in a
semantic sense, where a linear term is still consumed exactly once --
this is compounded by lazy evaluation driving a non-trivial mismatch
between syntactic and semantic linearity. Therefore, linear linting
rejects various valid programs (with regard to linearity) after
desugaring. The current solution to this is to effectively disable the
linear linter, since otherwise disabling optimisations can incur significant
performance costs. However, we believe that GHC's transformations are
correct, and it is the linear linter and its underlying type system
that cannot sufficiently accommodate the
resulting programs.
% diverse problems with linearity spring up when we get past the desugarer and
% the
% linter rejects valid programs which is quite undesirable.

@@

Despite not being accepted by the surface-level language, linear programs using
lets occur naturally in Core due to optimising transformations that create let
bindings. In a similar fashion, programs which violate syntactic linearity
for other reasons other than let bindings are produced by Core transformations.

@@

The key idea to computing the usage environment of multiple variables that use
themselves in their definition body is to perform the computation in two separate passes. First,
calculate a naive environment by emitting free variables multiplicities as usual
while treating the recursive-let bound variables as free ones. Second, pass the
variables names and their corresponding naive environment as input to
algorithm~\ref{computeRecUsages} to get a final usage environment. Intuitively,
the algorithm, for each recursive binder, iterates over all (initially naive) usage
environments and substitutes the recursive binder by the corresponding usage
environment, scaled up by the amount of times that recursive binder is used in
the environment being updated.

### 2023-01-23 report.tex
More on linear haskell

As an example, consider the following C-like program in which allocated memory
is freed twice. We know this to be the dreaded double-free error which will
crash our programs and crush our dreams. Regardless, a C-like type system will
accept this program without any issue.
\begin{code}
let p = malloc 4;
 in free p;
    free p;
\end{code}

@@

\begin{itemize}
    \item Linear types were only added to the language roughly \emph{31 years
        after} Haskell's inception, unlike Rust and Idris2 which were
        designed with linearity from the start. It is an especially difficult
        endeavour to add linear types to a well-established language with a
        large library ecosystem and many active users, rather than to develop
        the language from the ground up with linear types in mind, and the
        successful approach as implemented in GHC 9.0, the leading Haskell
        compiler, was based on Linear Haskell\cite{linearhaskell}, where a
        linear type system designed with retaining backwards-compatibility and
        allowing code reuse across linear and non-linear users of the same
        libraries in mind was described. We describe Linear Haskell in detail in
        section~\ref{sec:linear-haskell}.

@@

@@

\chapter{Background}

\section{Linear Types\label{linear-types}}

@@

Finally, the modality operator ($\bang$) constructs a resource that can be used
unrestrictedly. Its introduction rule ($\bang I$) states that to construct this
resource means to add a resource to the unrestricted context, which can then be
used freely. To construct an unrestricted value, however, the linear context
must be empty -- an unrestricted value can only be constructed if it does not
depend on any linear resource.
%
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
\]
%
To utilize an unrestricted value $M$, we must
bind it to $u$ with $\llet{\bang u = M}{N}$ and can then use it in $N$
unrestrictedly, because $u$ extends the unrestricted context rather than the
linear context as we have seen thus far.

@@

% TODO: Variable rules

\begin{figure}[h]
\[
  \begin{array}{c}
    \infer*[right=($\tensor I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\quad
    \infer*[right=($\tensor E$)]
    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Gamma ; \Delta , u{:}A \vdash M : B}
    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta \vdash N : B}
    {\Gamma ; \Delta \vdash M \with N : A \with B}
    \quad
    \infer*[right=($\with E_L$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
\quad
    \infer*[right=($\with E_R$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Gamma ; \Delta \vdash M : A}
    {\Gamma ; \Delta \vdash \textrm{inl}~M : A \oplus B}
\qquad
    \infer*[right=($\oplus I_R$)]
    {\Gamma ; \Delta \vdash M : B}
    {\Gamma ; \Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]
    \infer*[right=($\oplus E$)]
    {\Gamma ; \Delta \vdash M : A \oplus B \and \Gamma ; \Delta', w_1{:}A \vdash N_1 : C \and
    \Gamma ; \Delta', w_2{:}B \vdash N_2 : C}
    {\Gamma ; \Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Gamma ; \cdot \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Gamma ; \Delta \vdash M : \star \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
\]
\caption{Typing rules for a linearly-typed lambda calculus}
\label{fig:llcrules}
\end{figure}

@@

Haskell is a functional programming language defined by the Haskell
Report~\cite{jones1999haskell,marlow2010haskell} and whose \emph{de-facto}
implementation is GHC, the Glasgow Haskell Compiler\cite{GHC}. Haskell is a
\emph{lazy}, \emph{purely functional} language, i.e., functions cannot
have side effects or mutate data, and contrary to many programming languages,
arguments are \emph{not} evaluated when passed to functions, but rather are only
evaluated when its value is needed. The combination of purity and laziness is
unique to Haskell among mainstream programming languages.

@@

Another defining feature of Haskell is its powerful type system. Features such
as type classes, arbitrary-rank types, functional dependencies, GADTs, type
families and (more recently) even impredicative polymorphism\cite{10.1145/3408971} have become
commonplace in Haskell code, enforcing application level invariants and program
correctness through the types. For example, consider the definition of
\texttt{head} in the standard library, a function which takes the first element
of a list by pattern matching on the list constructors.

@@

If the list is empty, we terminate the program with an error, otherwise, we take
the first element. This function is unsafe -- our program might crash if we use
it on an invalid input.
\todo[inline]{When applied to the empty list, ...}

A common technique in Haskell is to use more expressive
types to assert properties about the values and get rid of the impossible cases
(e.g. a \texttt{NonEmpty} is a list that can not be empty).

\todo[inline]{Leveraging Haskell's more advanced features, we can .... }

A well liked motto is
"make invalid states unrepresentable". In this light, we will\todo{nao
  uses o futuro} introduce
Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
for which we can write a completely safe \texttt{head} function by expressing
that the size of the list must be at least one, at the type level.

\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a very popular\todo{nope} feature with
which users can define data types as they would common algebraic data
types, with the added ability to give explicit type signatures for the data
constructors where the result type may differ in the type parameters across
branches.
\todo{nao ha branches, é nos varios construtores. This allows for
  additional type invariants to be represented with GADTs through
  their type parameters, which restricts the use of specific
  constructors and their subsequent deconstruction through pattern
  matching}

Pattern matching against GADTs can introduce local type refinements,
that is, pattern matching against certain data constructors will refine the type
information used for typechecking individual case alternatives.
\todo{acho que a outra frase que escrevi e melhor que esta / complementa}
We develop
the length-indexed lists example without discussing the type system and type
inference details of GADTs which we later explore in~\ref{related-work-gadts}.

@@

The length-indexed list is defined inductively as either an empty list \emph{of
size zero}, or the construction of a list by appending a new element to an
existing list \emph{of size $m$} whose final size is $m+1$ (\texttt{S m}). The
list \texttt{Cons~'a'~(Cons~'b'~Nil)} has type \texttt{Vec~(S~(S~Z))~Char}
because \texttt{Nil~::~Vec~Z~Char} and \texttt{Cons~'a'~Nil~::~Vec~(S~Z)~Char}.
GADTs make possible different data constructors being parametrized over
different type parameters as we do with \texttt{Vec}'s size parameter being
different depending on the constructor that constructs the list.

To define the safe \texttt{head} function, we must specify the type of the input
list taking into account that the size must not be zero. To that effect, the
function takes as input a \texttt{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, which makes it impossible to pass a list of type
\texttt{Vec Z a} to \texttt{head}.

@@

Pattern matching on the \texttt{Nil} constructor would result in an innacessible
pattern warning because it is impossible to pass a list that would match
\texttt{Nil} -- it would entail the argument list has size zero; however, it's
impossible to call \texttt{head} on a list of size zero!

Linear types, which allow for a finer control over the usage of certain resources at the type
level, are another example of an extension to Haskell's type system
%, besides the aforementioned type-level features
that makes it even more expressive.

@@

Introduction of linear types to Haskell. How do they work, what's the syntax?
Multiplicities, examples. Monads, multiplicity polymorphism, multiplicity
polymoprhism applied to monadic bind to define resource-aware sequencing of
operations. Linear haskell relates systemFC with the linear lambda calculus?

@@

\chapter{Related Work}

@@

\section{System FC}

@@

\section{Linear Haskell}

@@

\section{OutsideIn(X)\label{related-work-gadts}}

@@

\section{Rust}

@@

\section{Introduction}

Since the publication of Linear Haskell~\cite{linearhaskell} and its release in GHC 9.0,
Haskell's type system supports linearity annotations in functions -- bringing
linear types to a mainstream, pure, and lazy functional language. Concretely,
function types can be annotated with a multiplicity (a multiplicity of One
requires the argument to be consumed exactly once, a multiplicity of Many allows
the argument to be consumed unrestrictedly, i.e., zero or more times). A
function is linear if it consumes its arguments exactly once when it itself is
consumed exactly once.
% TODO: for some definition of \emph{consume} that I should revise here.

>>>>

As an example, consider the following C-like program in which allocated memory
is freed twice. We know this to be the dreaded double-free error which will
crash our programs and crush our dreams. Regardless, a C-like type system will
accept this program without any issue.
\begin{code}
let p = malloc(4);
 in free(p);
    free(p);
\end{code}

@@

\begin{itemize}
    \item Linear types were only added to the language roughly \emph{31 years
        after} Haskell's inception, unlike Rust and Idris2 which were
        designed with linearity from the start. It is an especially difficult
        endeavour to add linear types to a well-established language with a
        large library ecosystem and many active users, rather than to develop
        the language from the ground up with linear types in mind, and the
        successful approach as implemented in GHC 9.0, the leading Haskell
        compiler, was based on Linear Haskell\cite{cite:linearhaskell}, where a
        linear type system designed with retaining backwards-compatibility and
        allowing code reuse across linear and non-linear users of the same
        libraries in mind was described. We describe Linear Haskell in detail in
        section~\ref{sec:linear-haskell}.

@@

% TODO: Rewrite this whole second part of the introduction.
% It needs to be much clearer

@@

\chapter{Background and Related Work}

\section{Linear Types\label{sec:linear-types}}

@@

Finally, the "of-course" operator ($\bang$) can be used to construct a resource that can be
used unrestrictedly ($\bang M$). Its introduction rule ($\bang I$) states that
to construct this resource means to add a resource to the unrestricted context,
which can then be used freely. To construct an unrestricted value, however, the
linear context \emph{must be empty} -- an unrestricted value can only be constructed if
it does not depend on any linear resource.
%
\[
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
\]
%
To utilize an unrestricted value $M$, we must
bind it to $u$ with $\llet{\bang u = M}{N}$ and can then use it in $N$
unrestrictedly, because $u$ extends the unrestricted context rather than the
linear context as we have seen thus far.

@@

\begin{figure}[h]
\[
  \begin{array}{c}
    \infer*[right=($u$)]
    { }
    {\Gamma ; u{:}A \vdash u : A}
\qquad
    \infer*[right=($u$)]
    { }
    {\Gamma , u{:}A ; \cdot \vdash u : A}
\\[1em]
    \infer*[right=($\tensor I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta , \Delta' \vdash (M \tensor N) : A \tensor B}
\quad
    \infer*[right=($\tensor E$)]
    {\Gamma ; \Delta \vdash M : A \tensor B \and \Gamma ; \Delta', u{:}A, v{:}B \vdash N : C}
    {\Gamma ; \Delta , \Delta' \vdash \llet{u \tensor v}{N} : C}
\\[1em]
    \infer*[right=($\lolli I$)]
    {\Gamma ; \Delta , u{:}A \vdash M : B}
    {\Gamma ; \Delta \vdash \lambda u. M : A \lolli B}
\qquad
    \infer*[right=($\lolli E$)]
    {\Gamma ; \Delta \vdash M : A \lolli B \and \Gamma ; \Delta' \vdash N : A}
    {\Gamma ; \Delta, \Delta' \vdash M~N : B}
\\[1em]
    \infer*[right=($\with I$)]
    {\Gamma ; \Delta \vdash M : A \and \Gamma ; \Delta \vdash N : B}
    {\Gamma ; \Delta \vdash M \with N : A \with B}
    \quad
    \infer*[right=($\with E_L$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{fst}~M : A}
\quad
    \infer*[right=($\with E_R$)]
    {\Gamma ; \Delta \vdash M : A \with B}
    {\Gamma ; \Delta \vdash \textrm{snd}~M : B}
\\[1em]
    \infer*[right=($\oplus I_L$)]
    {\Gamma ; \Delta \vdash M : A}
    {\Gamma ; \Delta \vdash \textrm{inl}~M : A \oplus B}
\qquad
    \infer*[right=($\oplus I_R$)]
    {\Gamma ; \Delta \vdash M : B}
    {\Gamma ; \Delta \vdash \textrm{inr}~M : A \oplus B}
\\[1em]
    \infer*[right=($\oplus E$)]
    {\Gamma ; \Delta \vdash M : A \oplus B \and \Gamma ; \Delta', w_1{:}A \vdash N_1 : C \and
    \Gamma ; \Delta', w_2{:}B \vdash N_2 : C}
    {\Gamma ; \Delta, \Delta' \vdash \ccase{M}{\textrm{inl}~w_1 \to N_1 \mid \textrm{inr}~w_2 \to N_2} : C}
    \\[1em]
    \infer*[right=($\star I$)]
    { }
    {\Gamma ; \cdot \vdash \star : \star}
\quad    \infer*[right=($\star E$)]
    {\Gamma ; \Delta \vdash M : \star \and \Gamma ; \Delta' \vdash N : B}
    {\Gamma ; \Delta, \Delta' \vdash \llet{\star = M}{N} : B}
\\[1em]
    \infer*[right=($\bang I$)]
    {\Gamma ; \cdot \vdash M : A}
    {\Gamma ; \cdot \vdash !M : !A}
\qquad    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta \vdash M : \bang A \and \Gamma, u{:}A ; \Delta' \vdash N : C}
    {\Gamma ; \Delta, \Delta' \vdash \llet{!u = M}{N} : C}
  \end{array}
\]
\caption{Typing rules for a linearly-typed lambda calculus}
\label{fig:llcrules}
\end{figure}

@@

Haskell is a functional programming language defined by the Haskell
Report~\cite{jones1999haskell,marlow2010haskell} and whose \emph{de-facto}
implementation is GHC, the Glasgow Haskell Compiler\cite{GHC}. Haskell is a
\emph{lazy}, \emph{purely functional} language, i.e., functions cannot
have side effects or mutate data, and, contrary to many programming languages,
arguments are \emph{not} evaluated when passed to functions, but rather are only
evaluated when its value is needed. The combination of purity and laziness is
unique to Haskell among mainstream programming languages.

@@

Another defining feature of Haskell is its powerful type system. In contrast to
most mainstream programming languages, such as OCaml and Java, Haskell supports a myriad of
advanced type level features, such as:
\begin{itemize}

    \item Multiple forms of advanced polymorphism: where languages with whole
        program type inference usually stick to Damas–Hindley–Milner type
        inference~\cite{DBLP:conf/popl/DamasM82}, Haskell goes much further with, e.g.,
        arbitrary-rank types~\cite{10.1017/S0956796806006034}, type-class
        polymorphism~\cite{10.1145/227699.227700}, levity polymorphism~\cite{10.1145/3062341.3062357},
        multiplicity polymorphism~\cite{cite:linearhaskell}, and, more recently, impredicative
        polymorphism~\cite{10.1145/3408971}.

    \item Type level computation by means of type classes and Haskell's type
        families~\cite{,,}, which permit a direct encoding of type-level
        functions resembling rewrite rules.

    \item Local equality constraints and existential types by using GADTs, we
        explain ahead in more detail. A design for first class existential types
        with bi-directional type inference in Haskell has been published
        in~\cite{10.1145/3473569}, despite not being yet implemented in GHC.

\end{itemize}
% Haskell's type families, supports abstraction through type classes, and GADTs
% which we explain in more detail ahead.  as type classes, arbitrary-rank types,
% functional dependencies, GADTs, type families and (more recently) even
% impredicative polymorphism\cite{}
These advanced features have become commonplace in Haskell code, enforcing
application level invariants and program correctness through the types. As an
example to work through this section while we introduce compile-time invariants
with GADTs, consider the definition of \texttt{head} in the standard library, a
function which takes the first element of a list by pattern matching on the list
constructors.

@@

When applied to the empty list, \texttt{head} terminates the program with an
error. This function is unsafe -- our program might crash if we use it on an
invalid input.
%
Leveraging Haskell's more advanced features, we can use more expressive types to
assert properties about the values and get rid of the invalid cases (e.g. we
could define a \texttt{NonEmpty} type to model a list that can not be empty).
%
A well liked motto is "make invalid states unrepresentable". In this light, we
introduce Generalized Algebraic Data Types (GADTs) and create a list type
indexed by size for which we can write a completely safe \texttt{head} function
by expressing that the size of the list must be at least one, at the type level.

\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are an advanced Haskell feature
which allows users to define data types as they would common algebraic
data types, with the added abilitiy to give explicit type signatures for the
data constructors where the result type may differ in the type parameters across
branches.
\todo{nao ha branches, é nos varios construtores. This allows for
  additional type invariants to be represented with GADTs through
  their type parameters, which restricts the use of specific
  constructors and their subsequent deconstruction through pattern
  matching}
% TODO: Rewrite this bit below
Pattern matching against GADTs can introduce local type refinements, that is,
pattern matching against certain data constructors will refine the type
information used for typechecking individual case alternatives. We develop the
length-indexed lists example without discussing the type system and type
inference details of GADTs which we later explore in~\ref{related-work-gadts}.

@@

The length-indexed list is defined inductively as either an empty list \emph{of
size zero}, or the construction of a list by appending a new element to an
existing list \emph{of size $m$} whose final size is $m+1$ (\texttt{S m}). The
list \texttt{Cons~'a'~(Cons~'b'~Nil)} has type \texttt{Vec~(S~(S~Z))~Char}
because \texttt{Nil} has type \texttt{Vec~Z~Char} and \texttt{Cons~'a'~Nil} has
type \texttt{Vec~(S~Z)~Char}.
GADTs make possible different data constructors being parametrized over
different type parameters as we do with \texttt{Vec}'s size parameter being
different depending on the constructor that constructs the list.

To define the safe \texttt{head} function, we must specify the type of the input
list taking into account that the size must not be zero. To that effect, the
function takes as input a \texttt{Vec (S n) a}, that is, a vector with size
(n+1) for all possible n's, making a call to \texttt{head} on a list of type
\texttt{Vec Z a} (an empty list) is a compile-time type error.

@@

Pattern matching on the \texttt{Nil} constructor is not needed, despite it being
indeed a constructor of \texttt{Vec}. The input type just doesn't match the type
of \texttt{Nil} ($\texttt{S~n} \neq \texttt{Z}$), so the corresponding pattern
case alternative would be innacessible because the typechecker will not allow
calling \texttt{head} on \texttt{Nil} (once again, its type, \texttt{Vec~Z~a},
does not match the input type of \texttt{head}, \texttt{Vec~(S~n)~a}).

In practice, the idea of using more expressive types to enforce invariants at
compile time, that is illustrated by this simple example, can be taken much
further, e.g., to implement type-safe artificial neural
networks\cite{https://blog.jle.im/entry/practical-dependent-types-in-haskell-1.html},
enforce size compatibility in operations between matrices and vectors\cite{}, to
implement red-black trees guaranteeing its invariants at compile-time, or to
implement a material system in a game engine\cite{}.
% TODO: cite: https://blog.jle.im/entry/practical-dependent-types-in-haskell-1.html
% TODO: haskell-servant?

Linear types are, similarly, an extension to Haskell's type system that makes it
even more expressive, by providing a finer control over the usage of certain
resources at the type level.
%, besides the aforementioned type-level features

@@

% Introduction of linear types to Haskell. How do they work, what's the syntax?
% Multiplicities, examples. Monads, multiplicity polymorphism, multiplicity
% polymoprhism applied to monadic bind to define resource-aware sequencing of
% operations. Linear haskell relates systemFC with the linear lambda calculus?

The introduction of linear types to Haskell's type system is originally
described in Linear Haskell\cite{cite:linearhaskell}. While
in~\ref{sec:related-work-linear-haskell} we discuss the reasoning and design
choices behind retrofitting linear types to Haskell, here we focus on
linear types solely as they exist in the language, and re-work the file
handle example seen in the previous section along the way.
% TODO: useless information but the term "Linear Haskell" is sometimes also used
% to refer to the existence of linear types in Haskell in general.

A linear function ($f :: A \lolli B$) guarantees that if ($f~x$) is consumed
exactly once, then the argument $x$ is consumed exactly once. The precise
definition of \emph{consuming a value} depends on the value as follows,
paraphrasing Linear Haskell~\cite{cite:linearhaskell}:

\begin{itemize}
    \item To consume a value of atomic base type (like \texttt{Int} or
        \texttt{Ptr}) exactly once, just evaluate it
    \item To consume a function value exactly once, apply it to one argument,
        and consume its result exactly once.
    \item To consume a value of an algebraic datatype exactly once,
        pattern-match on it, and consume all its linear components exactly once.
        For example, a linear pair (equivalent to $\tensor$) is consumed exactly
        once if pattern-matched on \emph{and} both the first and second element are
        consumed once.
\end{itemize}

In Haskell, linear types are introduced through \emph{linearity on the function
arrow}, rather than through \emph{linearity in the kinds}. In practice, this
means function types are annotated with a linearity that defines whether a
function argument must be consumed \emph{exactly once} or whether it can be
consumed \emph{unrestrictedly} (many times). Additionally, algebraic data types
constructor consume its arguments linearly by default, but the constructor can
be specified to receive unrestricted arguments which, in contrast, do not need
to be consumed exactly once when pattern matched on.
%
As an example, consider from the following functions the first, which doesn't
typecheck because it is a linear function (annotated with the linearity of
\texttt{1}) that consumes its argument more than once, and the second, which is
an unrestricted function (annoted with \texttt{Many}) that typechecks because
its type allows the argument to be consumed unrestrictedly.

\begin{code}
    f :: a %1 -> (a, a)
    f x = (x, x)
\end{code}
\begin{code}
    g :: a %Many -> (a, a)
    g x = (x, x)
\end{code}

The function annotated with the linearity annotation of \texttt{1} is equivalent
to the linear function type ($\lolli$) seen in the linear lambda
calculs~\ref{sec:linear-types}. To encode the multiplicative linear pair
($\tensor$) we must create a pair data type with two linear components. To
consume an algebraic data type is to consume all its linear components once, so
to consume said pair data type we need to consume both its linear components,
which successfully encodes the multiplicative pair elimination rule ($\tensor
E$). To construct said pair data type we must provide two linear elements, each
consuming some required resources to be constructed, which encodes the
multiplicative pair introduction rule ($\tensor I$).

\begin{code}
data MultPair a b where
    MkPair :: a %1 -> b %1 -> MultPair a b
\end{code}

The linearity annotations are a specialization of the more general so-called
\emph{multiplicity annotations}. A multiplicity of \texttt{1} entails that the
function argument must be consumed once, and a function annotated with it
($\to_1$) is called a linear function (also known as $\lolli$). A function with
multiplicity of \texttt{Many} ($\to_\omega$) allows the function argument to
be consumed unrestrictedly. Unrestricted functions are equivalent to the common
function type and, in fact, the usual function arrow ($\to$) implicitly has
multiplicity \texttt{Many}.  Multiplicities also smoothly allow for
\emph{multiplicity polymorphism}, which we describe below.

@@

\section{Related Work}

@@

\subsection{System FC}

@@

\subsection{Linear Haskell\label{sec:related-work-linear-haskell}}

@@

Haskell, contrary to most programming languages with linear types or
linearity-based types (such as ownership types), has existed for 31 years of its
life \emph{without} linear types. As such, the introduction of linear types to
Haskell comes with added challenges that can't exist in linearly-typed languages
that were designed with linear types from the start:

\begin{itemize}
    \item Backwards compatibility. The addition of linear types shouldn't break
        all existing Haskell code.
    \item Code re-usability. The linearly-typed part of Haskell's ecosystem and
        its non-linearly-typed counterpart should fit in together and it must be
        possible to define functions readily usable by both sides
        simultaneously.
\end{itemize}

\begin{itemize}
    \item Forwards compatibility -- Haskell, despite being an
        industrial-strength language, is also a petri-dish for experimentation
        and innovation in the field of programming languages. Therefore, Linear
        Haskell takes care to accomodate possible future features, in
        particular, its design is forward compatible with affine and dependent
        types.
\end{itemize}

\subsection{OutsideIn(X)\label{related-work-gadts}}

@@

\subsection{Rust}

@@

\section{Summary}

Since the publication of Linear Haskell~\cite{cite:linearhaskell} and its release in GHC 9.0,
Haskell's type system supports linearity annotations in functions -- bringing
linear types to a mainstream, pure, and lazy functional language. Concretely,
function types can be annotated with a multiplicity (a multiplicity of One
requires the argument to be consumed exactly once, a multiplicity of Many allows
the argument to be consumed unrestrictedly, i.e., zero or more times). A
function is linear if it consumes its arguments exactly once when it itself is
consumed exactly once.
% TODO: for some definition of \emph{consume} that I should revise here.
